Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-08-31 analysis/paper/paper.Rmd
Rewrote the code for violin plot to use the more efficient facet wrap function

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}

#Make long version of Points df to prep for violin plots
Points_long <-
  Points %>%
    select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
      gather(key = "Element", value = "Value")

#Filter long df to focus on Mg, and generate a violin plot
Mg <-
  Points_long %>%
  filter(Element == "X12Mg") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="salmon1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X12Mg))))) +
  theme_bw() +
  ylim(0,5) +
  xlab("Mg") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Al, and generate a violin plot
Al <-
  Points_long %>%
  filter(Element == "X13Al") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="lightgoldenrod1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X13Al))))) +
  theme_bw() +
  ylim(0,10) +
  xlab("Al") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Si, and generate a violin plot
Si <-
  Points_long %>%
    filter(Element == "X14Si") %>%
      ggplot(aes(x=Element, y=Value)) +
      geom_violin(fill="lightgoldenrod1") +
      ggtitle(paste("n = ", length(which(!is.na(Points$X14Si))))) +
      theme_bw() +
      ylim(10,100) +
      xlab("Si") +
      theme(
        axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
        plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
        axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#Filter long df to focus on P, and generate a violin plot
P <-
  Points_long %>%
  filter(Element == "X15P") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="lightgoldenrod1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X15P))))) +
  theme_bw() +
  ylim(0,2) +
  xlab("P") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on S, and generate a violin plot
S <-
  Points_long %>%
  filter(Element == "X16S") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="lightgoldenrod1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X16S))))) +
  theme_bw() +
  ylim(0,2.5) +
  xlab("S") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on K, and generate a violin plot
K <-
  Points_long %>%
  filter(Element == "X19K") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="salmon1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X19K))))) +
  theme_bw() +
  ylim(0,10) +
  xlab("K") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Ca, and generate a violin plot
Ca <-
  Points_long %>%
  filter(Element == "X20Ca") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="salmon1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X20Ca))))) +
  theme_bw() +
  ylim(0,1) +
  xlab("Ca") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on V, and generate a violin plot
V <-
  Points_long %>%
  filter(Element == "X23V") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X23V))))) +
  theme_bw() +
  ylim(0,0.05) +
  xlab("V") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Mn, and generate a violin plot
Mn <-
  Points_long %>%
  filter(Element == "X25Mn") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X25Mn))))) +
  theme_bw() +
  ylim(0,0.75) +
  xlab("Mn") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Fe, and generate a violin plot
Fe <-
  Points_long %>%
  filter(Element == "X26Fe") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X26Fe))))) +
  theme_bw() +
  ylim(0,2) +
  xlab("Fe") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Zn, and generate a violin plot
Zn <-
  Points_long %>%
  filter(Element == "X30Zn") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X30Zn))))) +
  theme_bw() +
  ylim(0,0.05) +
  xlab("Zn") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Sr, and generate a violin plot
Sr <-
  Points_long %>%
  filter(Element == "X38Sr") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="salmon1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X38Sr))))) +
  theme_bw() +
  ylim(0,0.02) +
  xlab("Sr") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Y, and generate a violin plot
Y <-
  Points_long %>%
  filter(Element == "X39Y") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X39Y))))) +
  theme_bw() +
  ylim(0,0.01) +
  xlab("Y") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Zr, and generate a violin plot
Zr <-
  Points_long %>%
  filter(Element == "X40Zr") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="skyblue1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X40Zr))))) +
  theme_bw() +
  ylim(0,0.1) +
  xlab("Zr") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Ba, and generate a violin plot
Ba <-
  Points_long %>%
  filter(Element == "X56Ba") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="salmon1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X56Ba))))) +
  theme_bw() +
  ylim(0,0.5) +
  xlab("Ba") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#Filter long df to focus on Ti, and generate a violin plot
Ti <-
  Points_long %>%
  filter(Element == "X81Ti") %>%
  ggplot(aes(x=Element, y=Value)) +
  geom_violin(fill="lightgoldenrod1") +
  ggtitle(paste("n = ", length(which(!is.na(Points$X81Ti))))) +
  theme_bw() +
  ylim(0,1.5) +
  xlab("Ti") +
  theme(
    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())

#16 figures arranged in 4 rows and 4 columns
grid.arrange(Mg,Al,Si,P,S,K,Ca,V,Mn,Fe,Zn,Sr,Y,Zr,Ba,Ti,ncol=4)
```

>>>>

```{r xrf-violin-plot, fig.width=15, fig.height=10, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points %>%
  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
  gather(key = "Element", value = "Value")

#Remove initial X introduced by read.csv to column names
Points_long$Element <- substring(Points_long$Element, 2)

# Add column with elemental groups for visualisation
Points_long <-
  Points_long %>%
    mutate(Group = case_when(
      endsWith(Element, "Mg") ~ "Alkaline Earth",
      endsWith(Element, "Al") ~ "Basic metal",
      endsWith(Element, "Si") ~ "Semimetal",
      endsWith(Element, "P") ~ "Nonmetal",
      endsWith(Element, "S") ~ "Nonmetal",
      endsWith(Element, "K") ~ "Alkali Metal",
      endsWith(Element, "Ca") ~ "Alkaline Earth",
      endsWith(Element, "V") ~ "Transition Metal",
      endsWith(Element, "Mn") ~ "Transition Metal",
      endsWith(Element, "Fe") ~ "Transition Metal",
      endsWith(Element, "Zn") ~ "Transition Metal",
      endsWith(Element, "Sr") ~ "Alkaline Earth",
      endsWith(Element, "Y") ~ "Transition Metal",
      endsWith(Element, "Zr") ~ "Transition Metal",
      endsWith(Element, "Ba") ~ "Alkaline Earth",
      endsWith(Element, "Ti") ~ "Transition Metal",
    ))

#Prepare colors
colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")

#Lookup for element labeller
Elements <- c(
  '12Mg' = "Mg",
  '13Al' = "Al",
  '14Si' = "Si",
  '15P' = "P",
  '16S'= "S",
  '19K' = "K",
  '20Ca' ="Ca",
  '23V' = "V",
  '25Mn' = "Mn",
  '26Fe' = "Fe",
  '30Zn' = "Zn",
  '38Sr' = "Sr",
  '39Y' = "Y",
  '40Zr' = "Zr",
  '56Ba' = "Ba",
  '81Ti' = "Ti"
  )

#Violin plot of elemental content, trimmed, with fill based on elemental groups
Points_long %>%
  drop_na(Value) %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=colors) +
    theme_bw() +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```
